Local Type Inference for Polarised System F with Existentials

### 2023-10-11 oopsla24/tex/03-declarative-system.tex
Wip: declarative system

\paragraph{Invariance of Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants makes the system equivalent to \systemf, 
and thus, undecidable. 
However, \ruleref{\ottdruleDOneShiftDLabel} might be
relaxed to a covariant form, as we will discuss in \cref{todo}.

As discussed in \cref{todo}, 
the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
For convenience of representation, we compose the left-hand side rule and
the right-hand side rule into one, and use substitution $[[σ]]$ to 
represent instantiation.

The substitution application is defined 
in a standard way, avoiding capture of bound variables,
and preserving the variables that are out of the substitution domain.
The domain and the range of the substitutions are
specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.

@@

@@

@@

Now, we present selected properties of the declarative system,
which are important for the correctness of the algorithm.

\begin{lemma}
  Variables do not have proper subtypes and supertypes
\end{lemma}

\begin{lemma}
  Free Variable propagation: $[[Γ ⊢ iN ≤ iM]]$ 
  implies $[[fv(iN) ⊆ fv(iM)]]$.
\end{lemma}

\begin{lemma}
  Subtyping is Reflexive and Transitive, and is preserved by substitution
\end{lemma}

\begin{lemma}
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) ≈ nf(iM)]]$.
\end{lemma}

\begin{lemma}[Characterization of Supertypes]
\end{lemma}

>>>>

\paragraph{Quantifiers} As discussed in \cref{todo}, 
the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
For convenience of representation, we compose the left-hand side rule and
the right-hand side rule into one, and use substitution $[[σ]]$ to 
represent instantiation.

\paragraph{Substitutions} The substitution application is defined 
in a standard way, avoiding capture of bound variables,
and preserving the variables that are out of the substitution domain.
The domain and the range of the substitutions are
specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.

\paragraph{Invariance of the Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, \ruleref{\ottdruleDOneShiftDLabel} might be
relaxed to a covariant form, as we will discuss in \cref{todo}.

@@

The declarative type inference system is shown in \cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to have the type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.
To infer the resulting type of an application, the
applicative let rules refer to the \emph{Application typing} judgment.
It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which is read as ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to have the type $[[iM]]$''.

@@

\paragraph{Non-bidirectionality}
  Notice that all the rules in our system are inferring.

\paragraph{Typing up to equivalence}
  As discussed in \cref{todo}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  The application typing is defined up to equivalence as well,
  although it has no special rules for that.

\paragraph{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.
  The difference is that the former rule is for the 
  \emph{unannotated} let binder, and thus
  it requires the resulting type of application to be unique
  (up to equivalence), so that the type of the bound variable $[[x]]$
  is known before it is put into the context.
  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants (see \cref{todo}).
  In \cref{todo} we discuss how this restriction can be relaxed
  together with invariant shift subtyping.

\paragraph{Application of a polymorphic type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  taken out of thin air. The algorithm we present in \cref{todo} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

\paragraph{Application of an Arrow Type}
  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{todo}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

\paragraph{Annotations}
  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
  inferred type is a subtype of the annotation type.

@@

Now we present selected properties of the declarative system,
which are important for the correctness of the algorithm.

As we discussed in \cref{todo}, the equivalence induced by the subtyping is 
not trivial. To deal with it without modifying standard operations such as unification, 
we use normalization. This way, it is important for the normalization to be 
sound and complete \wrt the equivalence:

\begin{property}[Correctness of normalization]
  For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$.
\end{property}

Another property that we extensively use is that the subtyping is reflexive and transitive. 
Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.

\begin{property}[Subtyping is well-defined]
  The subtyping is a preorder and is preserved under substitutions. 
\end{property}

However, the greatest lower bound of two positive types is not always defined.
Let us consider a positive type $[[iP]]$,
and two different supertypes $[[iP1]]$ and $[[iP2]]$
(\ie  $[[· ⊢ iP1 ≥ iP]]$, $[[· ⊢ iP2 ≥ iP]]$, and $[[nf(iP1) ≠ nf(iP2)]]$.
Let us also consider an arbitrary negative type $[[iN]]$.
